---
title: "HW2-EDA with R - Part1: Plotting"
author: "misken"
date: "January 22, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. 
## When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.
Do this
## Step 3 - Set working directory and explore data folder

Set the working directory or create an R Studio project in the current folder. You'll notice that there is a folder named **data**.
Inside of it you'll find the data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.
You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(dplyr)
library(ggplot2)
```

### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_MI.csv** into a data frame named county_MI.

```{r read_countydata}
county_df <- read.csv("data/county.csv", header = TRUE)
county_MI <- read.csv("data/county_MI.csv", header = TRUE)
```

Check the structure of each data.frame and use the `summary` function to begin
exploring the data. Are there any NA values? What 
differences do you notice between the structure of the two data frames.

```{r}
str(county_df)
```

```{r}
str(county_MI)
```

The Michigan-only data has a column for peninsula, and the County, State,  
CensusId, State.Code, Region, and Division columns are in a different order.
The State, State.Code, Region, and Division columns all contain the same value.

```{r}
summary(county_df)
```

The full data set has NA values in Income, Income Error, 
and Child Poverty columns

```{r}
summary(county_MI)
```

The Michigan only dataset does not contain any NA values.


The county_MI data frame contains data for Michigan only and includes an
peninsula field. See **census_data_prep.R** to see how this field was
created. It involves scraping a table out of a web page.

Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.

### Problem 2 - Basic plots for Income

Create a histogram, boxplot, kernel density plot and violin plot for median
income for the entire US. 

Write a short paragraph describing the distribution of median incomes.

```{r}
g <- ggplot(data = county_df) 

g + geom_histogram(aes(x = Income))

g + geom_boxplot(aes(y = Income, x = 1)) 

g + geom_density(aes(x = Income), fill="grey50")

g + geom_violin(aes(y = Income, x = 1)) 

```

#Median Income Analysis
Nearly all counties have a median income between 25,000 and 75,000, 
though there are a few dozen outliers with larger median incomes up to 125,000. 
50% of counties have median incomes between 37,000 and 55,000. 


### Problem 3 - Histogram of total population

Create a histogram of total population for counties in the US.

```{r histo_totpop}
g + geom_histogram(aes(x = TotalPop))
```

Repeat the histogram about but make the following style changes to the plot.

* Add a title such as "Distribution of population for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Total Population"
* Change the colors of the bars to something other than the default black

```{r histo_stylechanges}
g + geom_histogram(aes(x = TotalPop), fill = "Blue") + 
  labs(title = "Distribution of population for US counties", 
       x ="Total Population", y = "Number of counties")
```

**Hacker Extra** - Also create a kernel density plot overlaid on a histogram of
Income. 

Repeat the histogram but only for counties with less than 2.5 million people.

```{r histo_lt2.5m}
ggplot(data = county_df, aes(x = Income, y = ..density..)) + 
  geom_histogram(fill = "Blue") + 
  labs(title = "Distribution of income for US counties", 
       x ="Income", y = "Number of counties") + 
  geom_density()

smaller <- county_df %>%
  filter(TotalPop < 2500000)

ggplot(data = smaller, aes(x = Income, y = ..density..)) + 
  geom_histogram(fill = "blue") + 
  labs(title = "Distribution of income for US counties", x ="Income", 
       y = "Number of counties") + 
  geom_density()
```


### Problem 4 - How do Michigan's peninsulas compare?

Create a grouped box plot showing Income by peninsula. Make sure
your graph has a title and good axis labels.

Do the peninsulas 
differ with respect to distribution of median income over the counties?


```{r bp_inc_by_peninsula}
ggplot(data = county_MI) + geom_boxplot(aes(y = Income, x = peninsula)) + 
  labs(title = "Distribution of income by peninsula", 
       y = "Median Income", x = "Peninsula")
```

The income distribution is quite different between the peninsulas. 
The 75th quantile of the Upper Peninsula is about the same as the 25th quantile 
of the Lower Peninsula. The highest incomes in the UP are roughly 45,000 whereas
in the Lower Peninsula, 25% of the median incomes are 50,000 or greater.
One should also take into account that there are fewer counties in the UP 
than in the LP. 

Create several more boxplots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

```{r more_boxplots}
ggplot(data = county_MI) + geom_boxplot(aes(y = Native, x = peninsula)) + 
  labs(title = "Percent of population that's Native American by peninsula", 
       y = "Native American %", x = "Peninsula")
```

The Upper Peninsula has a much greater percentage of Native American 
residents in some of its counties than than the Lower Peninusula. The
median percentage of the UP is greater than all but two counties in the
Lower Peninsula.


```{r}
ggplot(data = county_MI) + geom_boxplot(aes(y = IncomePerCap, x = peninsula)) + 
  labs(title = "Distribution of income per capita by peninsula", 
       y = "Income Per Person", x = "Peninsula")
```
Income Per Person is slightly more even than Median Income between the 
peninsulas. Here the median in the Lower Peninsula is equal equivalent
to the 75th percentile in the Upper Peninsula. 


```{r}
ggplot(data = county_MI) + geom_boxplot(aes(y = Poverty, x = peninsula)) + 
  labs(title = "Percent of population in poverty by peninsula", 
       y = "% in Poverty", x = "Peninsula")
```

The 75th percentiles of poverty percentage are nearly the same between
peninsulas. The Lower Peninsula has a wider spread, with outlier counties
at 6% and 30%.

```{r}
ggplot(data = county_MI) + geom_boxplot(aes(y = ChildPoverty, x = peninsula)) + 
  labs(title = "Percent of children in poverty by peninsula", 
       y = "% of Children in Poverty", x = "Peninsula")
```

The peninsulas are similar when it comes to the percent of children in 
poverty, althought the Lower Peninsula has a greater range. Sadly, all
counties in the UP have at least 15% of their children in poverty. Median
child poverty by peninsula is nearly the same, just under 25%.


```{r}
ggplot(data = county_MI) + geom_boxplot(aes(y = MeanCommute, x = peninsula)) + 
  labs(title = "Distribution of mean commute by peninsula", 
       y = "Commute time (mins)", x = "Peninsula")
```

I was surprised to see that mean commute is lower in the Upper Peninsula.
75% of counties in the Upper Peninsual have an average commute 
time of 21 minutes or less. 75% of counties in the Lower Peninsula have
an average commute time of 21 minutes or more. Perhaps this is because 
there are fewer big cities into which people have long commutes.



### Problem 5 - How do regions of the country compare?

In county_df you'll see that there's a Region field. Create a set of faceted 
histograms showing how some variable of interest to you compares by region. For 
example, you could explore how the regions compare with respect to the
distribution of MeanCommute time. Make sure your graph has a title and good axis
labels.

```{r faceted_histo}
# I created this column to show the proportion of women to men
county_df$WomenRatio <- county_df$Women/county_df$Men

ggplot(data = county_df) + 
  geom_histogram( aes(x = WomenRatio, y = ..density..), bins = 30) + 
  geom_density(aes(x = WomenRatio)) +
  facet_wrap( ~ Region) + 
  labs(title = "Distribution of women to men ration by region", 
       x = "Ratio of women to men", y = "Number of Counties") +
  xlim(0,1.5)
```
Though there are no data points as low as 0 or as high as 1.5, I 
changed the axis units in order to more clearly show where the 1.0 
mark was.In the Northeast and the South, most counties have more 
women than men. 
### Problem 6 - Basic bar chart

Create a bar chart showing the number of counties by Region in the US. Make sure
your graph has a title and good axis labels.

```{r bar_counties_by_region}
ggplot(data = county_df) + 
  geom_bar(aes(x = Region)) + 
  labs(title = "Number of Counties by Region", 
       y = "Number of Counties", x = "Region")
```

### Problem 7 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
Income and MeanCommute.

```{r xy_inc_commute}
ggplot(data = county_MI) + geom_point(aes(x = Income, y = MeanCommute))
```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula.

```{r xy_inc_commute_peninsula}
ggplot(data = county_MI) + 
  geom_point(aes(x = Income, y = MeanCommute, color = peninsula))
```

Your thoughts on the relationship between these variables?
In the Upper Peninsula, jobs with longer mean commutes have lower income.
This could represent areas where jobs are scarce, so you have to go farther
to find any work. There seems to be a lot more variability in the Lower Peninsula.

Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?

```{r xy_faceted_peninsula}
# Do this
ggplot(data = county_MI) + 
  geom_point(aes(x = MeanCommute, y = FamilyWork, color = peninsula))
```


Now use the full US data frame.

Create a scatter plot of TotalPop by Income.

```{r xy_pop_inc}
ggplot(data = county_df) + 
  geom_point(aes(x = Income, y = TotalPop))
```

The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.
```{r}
ggplot(data = county_df) + 
  geom_point(aes(x = log(Income), y = TotalPop))
ggplot(data = county_df) + 
  geom_point(aes(x = Income, y = log(TotalPop), color = Region))
ggplot(data = county_df) + 
  geom_point(aes(x = log(Income), y = log(TotalPop), color = Region))
```
When you have variables that have a very wide range, obvious patterns may not be visible 
with a normal scale. Putting the variables into a log scale, makes patterns more obvious.
In this case, putting both the scales into a log transform allows the greatest number
of data points to be seen in the graph. 

### Hacker Extra - More analysis

Do some additional plotting to look for interesting patterns or relationships
either in the full US dataset or the MI only dataset. In addition, try out
some of ggplots additional features such as themes, axis formatting, gridline
control, or others.

Do this